---
title: 'gpbalancer: Optimally Balanced Gaussian Process Propensity Score Estimation'
author: "Brian Vegetabile"
date: "11/17/2017"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# $\texttt{gpbalancer}$: an `R` Package for Optimally Balanced Gaussian Process Propensity Score Estimation

Reference forthcoming...


## Installing `gpbalancer`

The package `devtools` is required to install this `R` package from this Github repository.  Install this package first if it is not already installed.

```{r, echo=TRUE, eval=FALSE}
install.packages('devtools', dependencies = TRUE)
```

Once that package has been installed, use the following to install `gpbalancer`

```{r, echo=TRUE, eval=FALSE}
devtools::install_github('bvegetabile/gpbalancer')
```

Load the package to begin analysis!

```{r, echo=TRUE, eval=T}
library('gpbalancer')
```

## Example using `gpbalancer`



### Simulating data

Provided below is a simple simulation to explore the effectiveness of the optimally balanced Gaussian process propensity score. Consider a continuous covariate $X$ which is used to assign treatment, let

\[
X_i \sim N(0,1)
\]

Additionally, let the true propensity score be defined as follows,

\[
e(X_i) = Pr(T_i = 1 | X_i) = 0.9 \times \Phi(2 * X_i) + 0.05
\]

where $\Phi()$ is the cumulative distribution of the Normal Distribution.  Finally, we simulate treatment assignment such that,

\[
T_i | X_i \sim Bernoulli()
\]

The true propensity score and observed treatment assignments are shown below.


```{r, echo=TRUE, eval=T}
set.seed(201711)
n_obs <- 500
pretreatment_cov <- rnorm(n_obs)
prop_score <- 0.9 * pnorm(2*pretreatment_cov) + 0.05
treatment_assignment <- rbinom(n_obs, size = 1, prob = prop_score)
t_col <- rgb(0.5,0,0.5,0.5)
c_col <- rgb(0,0.5,0,0.5)
```


### Visualizing the Propensity Score & Covariate Imbalance
```{r, echo=TRUE, eval=T, fig.width=14}
par(mfrow=c(1,2))
plot(pretreatment_cov, prop_score, 
     xlim=range(pretreatment_cov), 
     ylim=c(0,1),
     pch=19, col=rgb(0,0,0,0.5),
     xlab='Pretreatment Covariate',
     ylab='Probability of Treatment',
     main='Propensity Score & Observed Treatment Assignments')
points(pretreatment_cov[as.logical(treatment_assignment)],
       treatment_assignment[as.logical(treatment_assignment)],
       pch=19, col=t_col)
points(pretreatment_cov[!as.logical(treatment_assignment)],
       treatment_assignment[!as.logical(treatment_assignment)],
       pch=19, col=c_col)
abline(h=c(0,1), lty=3)
legend('topleft', c('Treated', 'Control'), pch=19, col=c(t_col, c_col), bg='white')

plot(density(pretreatment_cov[!as.logical(treatment_assignment)], adjust = 1.5),
     xlim=range(pretreatment_cov), 
     type='l', lwd=3, col=c_col,
     xlab='Pretreatment Covariate',
     ylab='Density',
     main='Covariate Distributions Conditional Upon Treatment Type')
lines(density(pretreatment_cov[as.logical(treatment_assignment)], adjust = 1.5),
     lwd=3, col=t_col)
abline(h=c(0,1), lty=3)
legend('topleft', c('Treated', 'Control'), lty=1, lwd=3, col=c(t_col, c_col), bg='white')

```

